---
title: "Celtic_Sea_Status_Quo_Model"
author: "Jessica Westworth"
format: pdf
editor: visual
---

# Introduction

Creating a model representing the average state of the Celtic Sea from 2012 to 2024. Composed of 14 species: herring (*Clupea harengus*), cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*), whiting (*Merlangius merlangus*), blue whiting (*Micromesistius poutassou*), hake (*Merluccius merluccius*), monkfish (Lophius piscatorius), horse mackerel (*Trachurus trachurus*), mackerel (*Scomber scombrus*), plaice (*Pleuronectes platessa*), megrim (*Lepidorhombus whiffiagonis*), sole (*Solea solea*), boarfish (*Capros aper*), and red gurnard (Chelidonicthys cuculus).

## Loading in Repository and Libraries

```{r}
#| warning: false
#| message: false
#| results: 'hide'

library(mizer)
library(mizerEcopath)
library(dplyr)
library(here)
```

## Load in Data

Data includes: life_history parameters (l_mat and age_mat calculated from DATRAS data; w_max and LWR from Fishbase); landing distributions form STECF; survey size distribution and age at length distribution data from DATRAS; Stomach contents data; Biomass estimates, yield estimates, interaction matrix, and diet matrix from Lauria (2016).

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#Load life history data
data(life_histories)

#Load in ecopath data
data(ecopath_params)

#Diet matrix
data(diet)

#Deaths from fishing
data(fishing_death)

#Catch Distribution
data(catch_distributions)

#DATRAS size distribution
data(survey_length_distributions)

#DATRAS age at length distribution
data(cs_age_size)
boar<-readRDS("/Users/jessicawestworth/Desktop/boarfish_struct_like_DATRAS.rds")
cs_age_size<-rbind(cs_age_size,boar)

#determining l_max comparing landings max length versus fishbase max length recorded value
observed_max <- catch_distributions %>%
     group_by(species) %>%
     summarise(
         obs_len = if (all(is.na(length + dl))) NA_real_ else max(length + dl, na.rm = TRUE)
     )
life_histories <- life_history_fishbase %>%
    left_join(observed_max, by = "species")%>%
    mutate(
        l_max = pmax(obs_len, l_max))%>%
    select(-obs_len)

#combined survey fishing and STECF fishing
all_fishing<-rbind(catch_distributions,survey_length_distributions)
```

## Species Parameters

Assign species parameters from life history data

```{r}
#calculate w_mat and w_max and set n,p,d, and alpha parameter
sp_params<-life_histories%>%
    mutate(w_mat=a*l_mat^b,
           w_max=a*l_max^b,
           n=0.7,
           p=0.7,
           d=-0.3,
           alpha=0.7,
           biomass_cutoff=w_mat)

species_to_groups <- list(
    Herring          = "Herring",
    Cod              = c("Cod", "Juvenile cod"),
    Haddock          = c("Haddock","Juvenile haddock"),
    Whiting          = c("Whiting","Juvenile whiting"),
    `Blue whiting`   = c("Blue whiting","Juvenile blue whiting"),
    Hake             = c("Hake","Juvenile hake"),
    Monkfish         = c("Monkfish","Juvenile monkfish"),
    `Horse mackerel` = "Horse mackerel",
    Mackerel         = "Mackerel",
    Plaice           = c("Plaice","Juvenile plaice"),
    Megrim           = c("Megrim","Juvenile megrim"),
    Sole             = "Sole",
    Boarfish         = "Boarfish",
    `Red gurnard`      = "Red gurnard"
)

sp <- sp_params
sp <- addEcopathParams(sp, ecopath_params, species_to_groups)

adult_map <- tibble::tibble(
  species      = names(species_to_groups),
  adult_stanza = vapply(species_to_groups, function(x) x[[1]], character(1))
)

adult_biomass <- ecopath_params %>%
  group_by(Group.name) %>%
  summarise(
    biomass_observed = sum(Biomass..t.km.2., na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(adult_stanza = Group.name)

sp <- sp %>%
  select(-any_of("biomass_observed")) %>%  # drop if present from a previous run
  left_join(adult_map, by = "species") %>%
  left_join(adult_biomass, by = "adult_stanza") %>%
  mutate(
    biomass_observed = coalesce(biomass_observed, 0),
    biomass_cutoff   = w_mat
  ) %>%
  select(-adult_stanza)

sp$biomass_cutoff[c(1,8,9,12,13,14)]<-1

sp<-sp[c(5,13,2,3,6,1,8,9,11,7,10,14,12,4),]
```

## Build Non-Interacting Model

```{r}
model_non_int <- newAllometricParams(sp, no_w = 200)

## Add gear params
model_non_int <- addCatch(model_non_int, landings= catch_distributions,
                              fishing_dead_biomass=fishing_death,
                              survey= survey_length_distributions,step=2)

# Steady model
model_non_int <- model_non_int %>%
    steadySingleSpecies() %>%
    matchBiomasses() %>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()

params<-model_non_int
sps<-species_params(model_non_int)
initial_Eiw<-array(dim=length(sps$species))
for(i in 1:length(sps$species)){
    initial_Eiw[i]<-getExtEncounter(params)[i,1]/(w(params)[1]^sp[i,]$n)}
sps$Eiw<-initial_Eiw
species_params(model_non_int)<-sps

#set the spawning time, spawning concentration to literature values.
spawning_mort<-tuneEcopath(
    model_non_int, 
    catch = all_fishing,
    age_at_length = cs_age_size)

#saveRDS(spawning_mort,"/Users/jessicawestworth/Desktop/spawning_mort.rds")
```

# Add Diets and Predation Kernel

```{r}
#| eval: false
# Stomach-fit parameters (keep separate; align by scientific name), cannot do right now because the new species are not in the processed file right now so... yeah...
data(stomach_data_fit)
stf<-stomach_data_fit
sp  <- species_params(spawning_mort)
idx <- match(sp$Scientific_name, stf$species)  # align order to species_params

lambda <- 2
sp$pred_kernel_type <- "power_law"
sp$kernel_exp <- stf$alpha[idx] + 4/3 - lambda
sp$kernel_l_l <- stf$ll[idx]
sp$kernel_u_l <- stf$ul[idx]
sp$kernel_l_r <- stf$lr[idx]
sp$kernel_u_r <- stf$ur[idx]

species_params(spawning_mort) <- sp

#Adjust diet matrix for integration into model using reduceEcopathDiet
dm<-reduceEcopathDiet(
    species_params = species_params(spawning_mort),
    ecopath_diet   = diet,
    ecopath_params = ecopath_params
)
```

# Continuous Re-tuning

```{r}
#| eval: false
p<-spawning_mort

#Tune model in the tuning app
non_int_tuned_model<-tuneEcopath(
    p, 
    catch = all_fishing,
    age_at_length = cs_age_size,
    diet = dm)

#Set Diffusion matrix as tuned app does not save into the model currently
non_int_tuned_model<-setDiffusion(non_int_tuned_model)

#Retrieve tuned model
data(non_int_tuned_model)
```

# Prepare Model for Projection

```{r}
#| eval: false
#ensure model is in a steady state
s<-mizer::steadySingleSpecies(non_int_tuned_model)
#set the feeding levels
g<-setFeedingLevels(params=s, f=0.6, f_c=0.2)
#ensure model is in a steady state
f<-mizer::steadySingleSpecies(g)

params <- f
dm <- dm
params <- validParams(params)

resource_params(params)$w_pp_cutoff <- 1

ps <- params |>
    mizer::steadySingleSpecies() |>
    alignResource() |>
    setResourceInteraction(resource_dynamics = "resource_semichemostat") |>
    matchDiet(dm) |>
    setBevertonHolt(reproduction_level = 0.8)

status_quo_interacting_model<-ps

#Retrieve previously tuned interacting model
data(status_quo_interacting_model)
```

# BH Model Projection Example

```{r}
#| eval: false
# Initialize the blender
# Let's say we want to transition over 20 years with a harvest intensity of 0.15, but then we want to keep fishing at BH till true steady state is reached.
t_duration <-350
og_sim_steady_in<-100
t_blend <- 200
c_intensity <- 0

my_blender <- make_blended_ssBH_FMort(t_max_blend = t_blend, 
                                      target_c = c_intensity,
                                      t_steady = og_sim_steady_in,
                                      alpha_max = 0.2)

# Inject function into params
sim_BH_start<- setRateFunction(ps, "FMort", "my_blender")

# Run the simulation
# Use effort = 1; the blender handles the internal weights
sim_BH_4 <- project(sim_BH_start, t_max = t_duration, effort = 1)

#plotBiomasses overtime
plotlyBiomass(sim_BH_4)
```
